Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-08-31 Tex/Paper.tex
Edits after meeting with Erik

\title{Emittance preservation of an electron bunch in a loaded quasi-linear plasma wakefield}

@@

\begin{abstract}
We investigate beam loading and emittance preservation for a high-charge electron beam being accelerated in quasi-linear
plasma wakefield driven by a short proton beam. The structure of the wakefield is similar to that of a long, modulated
proton beam. By selecting transverse and longitudinal electron beam parameters in order to appropriately load the wake,
we show that the bulk of the electron beam can be accelerated without significant emittance growth.
\end{abstract}

@@

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativist beam travelling through a plasma
excites a strong longitudinal e-field that can be loaded by a trailing witness beam. With optimal beam loading, increase
in energy spread can be kept to a minimum. Acceleration of an electron bunch by an electron drive bunch has been
demonstrated experimentally \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} in the past. AWAKE at CERN is a proof
of concept proton driven plasma wakefield accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$ the transverse wakefields cause emittance growth of the accelerated bunch. A
finite length bunch will also see a transversely varying accelerating field causing increasing energy spread
\cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, the drive beam sees a constant transverse field
due to a bubble forming in the plasma behind the drive beam. The bubble is formed by the transverse oscillations of the
plasma electrons in the wake of the drive bunch, forming a sheet around an evacuated area. The much heavier plasma ions
have no significant transverse movement within the time frame of the plasma electron oscillation as these are
proportional to $sqrt{m}$. The ions form an ion channel creating a focusing force that varies linearly with radius. This
produces an angularly symmetric focusing force \cite{lu:2006-1, lu:2006}. This effect also avoids the issue of
increasing energy spread due to transverse beam loading. Increase in energy spread is still dependent on beam loading
$E_{z}$. Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
\cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

The energy carried by electron drive bunches used in previous experiments is typically small, on the order of
$100\unit{J}$, and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. De-phasing of the
witness bunch relative to the drive bunch, caused by loss of energy in the drive bunch and gain of energy in the witness
bunch, limits the effective length of a plasma accelerator stage. Proton drive beams available at CERN carry
significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for longer accelerator
stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications.
This issue is resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of
the buckets in the modulated structure. This self-modulation instability is caused by the transverse fields generated by
the beam acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation
frequency is close to that of the plasma, and leaves a train of proton bunches along the beam axis with a surrounding
halo.

% ******************************************************************************************************************** %
\subsection[\label{S:I:AWAKE}]{The AWAKE Experiment}
% ******************************************************************************************************************** %

The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver. The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell. The rubidium vapour is
ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$. The nominal
plasma density of the AWAKE experiment is $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron bunch. The study presented here is for Run 2, which aims to demonstrate acceleration of a
short electron bunch to high energy with a minimal increase in emittance and energy spread.

@@

The preliminary design of the second run proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without
the electron beam present. The electron witness beam will be injected into the modulated proton beam before stage two,
where acceleration will occur. As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
keep this as short as possible \cite{adli:2016}.

\vspace{1em}
\noindent\textbf{From Patric, expand and add this:}\\
Then I would propose to state that the SM does not create a focusing force increasing linearly with radius because not
all e- are evacuated. The state that because the structure is small, a  bunch with significant charge and reasonable
current is not much shorter than the structure size and thus large energy spread is expected.

@@

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations. The proton beam was pre-modulated by a clipped cosine function to its
longitudinal density profile, with a period matching $\lambda_{pe}$. Its transverse profile was kept as a Gaussian with
$\sigma_{r} = k_{pe}^{-1} = 200\unit{\mu m}$. The length of the simulation beam was limited to $26\cdot\lambda_{pe}$,
and the electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed several parameter
scans with this setup, testing for the the maximum beam charge that could be accelerated with a high energy gain and low
energy spread \cite{adli:2016, berglyd_olsen:2016}.

In order to evaluate the quality of the beam, we also needed to study emittance evolution. However, full PIC codes like
Osiris are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect'' \cite{godfrey:1974}.
This is a know issue with the Yee EMF solver which causes the phase velocity of electromagnetic fields to be lower than
$c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing the Yee solver with a
solver designed by Lehe \cite{lehe:2013}. Our simulations showed that this numerical effect is still prominent in the
high density regions of the electron beam, making it difficult to distinguish emittance growth from the physics from
that originating from numerical error.

Quasi-static PIC codes do not suffer this problem, so in order to study the emittance evolution of the beam we instead
turned to the recently released open source version QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D PIC
code \cite{huang:2006, an:2013}.

@@

In the QuickPIC simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we
expect to see from the self-modulated SPS beam in AWAKE Run 2. The baseline AWAKE drive beam contains $3\nexp{11}$
protons \cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded
in the middle of the unmodulated bunch. In addition. a significant portion of the protons are lost during
self-modulation. The needed gap between the two plasma cells also contributes to a decreased density of protons near the
axis, and some of the field is drained by other protons seeing an accelerating phase within the drive beam itself. In
total, the expected peak $E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to $500-600\funit{MV}{m}$ in
the second cell \cite{awake_collaboration:2016}. A single proton bunch of $1.46\nexp{10}$, or $2.34\unit{nC}$ and
$7\unit{kA}$ was able to drive an $E_{z}$ field of a comparable magnitude in our single drive bunch setup.

The single bunch approach also has the benefit of reducing simulation time for the 3D simulations. Our aim was to create
a stable environment for the witness beam. We are not considering the evolution of the proton beam in this test setup,
so to prevent the proton beam from evolving radially, we increased the drive beam particle mass by a factor of
$1\mexp{6}$.

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam in
reference simulations \cite{awake_collaboration:2016}. This condition is replicated in out single bunch case. The peak
density of the drive beam is $0.83\cdot n_{0}$ producing a quasi-linear condition behind it. The single bunch setup uses
the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size $\sigma_{r} = 200\unit{\mu m}$
throughout the plasma cell. The length of the drive bunch $\sigma_{z} = 40\unit{\mu m}$.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. AWAKE baseline energy for the long witness beam for Run 1 is $W_{eb} = 10-20\unit{MeV}$
\cite{gschwendtner:2016}. In Run 2 the experiment will likely require a higher initial energy, but a
$W_{eb} \approx 50\unit{MeV}$ is sufficient to minimise de-phasing as the drift $\Delta\xi \propto \gamma^{-2}$
\cite{berglyd_olsen:2015}. For this case we have eliminated this source of de-phasing entirely in order to reduce the
number of variables.

Earlier simulations with witness beam radius $\sigma_{r} = 100\unit{\mu m}$ caused the beam to undergo rapid transverse
evolution as the beam entered the plasma region, reducing its $\sigma_{r}$ from $100\unit{\mu m}$ to $< 10\unit{\mu m}$
\cite{berglyd_olsen:2016}. This was primarily due to a mismatch between beam emittance and the plasma density. This
relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength.

In these simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{r}$ of $5.25\unit{\mu m}$. The relatively small size of the witness beam compared to the proton beam put some
restrictions on the transverse resolution of the witness beam. The transverse grid cell resolution was set to
$1.17\unit{\mu m}$.

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent wakefields throughout the plasma stage. The $E_{z}$-field generated by the proton drive bunch is seen as
the blue line in figure \ref{Fig:BeamLoading}, shown with and without the electron beam present. With a proton beam
density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near non-linear \cite{rosenzweig:2010}. There is
some depletion of plasma electrons, but not enough to form a bubble. The dashed green line in the lower part of figure
\ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of $67\%$, close to what we see in full
scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

The witness bunch generates its own wakefield, which in the accelerating phase partially cancels out the $E_{z}$ field
generated by the drive bunch. For a finite length bunch this causes the particles in the tail of the witness bunch to be
accelerated less than those at the front resulting in higher energy spread in exchange for higher energy transfer
\cite{van_der_meer:1985}. Since $E_{z} \propto \rho_{b}$, a beam profile $\rho_{b}(\xi)$ that perfectly flattens the
accelerating field within the witness beam will reduces energy spread to zero. The ideal shape is triangular or
trapezoidal with the peak density in the direction of the beam velocity \cite{katsouleas:1987, tzoufras:2009}.

@@

For Gaussian beams a reasonable flat field, and consequently low energy spread, can be achieved by controlling the
charge density to achieve a similar effect. This, however, puts the front of the witness bunch outside of the ideal
region while the middle bulk of the beam has a trapezoidal-like shape. In our case, with a matched beam to the plasma
density defined by equation \ref{EQ:Matched}, we are limited on how much total charge we can accelerate without reaching
charge densities that will overload the field. The test case shown in figures \ref{Fig:BeamLoading} and
\ref{Fig:PlasmaDenTWake} has a total charge of $100\unit{pC}$ with a $\sigma_{r}=5.23\unit{\mu m}$ matching an initial
normalised emittance $\epsilon_{0} = 2\unit{\mu m}$.

@@

Our configuration also shows some robustness to small electron bunch transverse offsets as the bubble generated by the
witness bunch wakefield follows the bunch itself. The head of the bunch does not benefit from this effect, and we see
some defocusing in this region. This effect is likely to be greater for larger offsets as the bunch oscillates around
the axis of the drive bunch wakefield (see right hand plots of figure \ref{Fig:BeamFilament}). However, since the proton
wakefield creates a quasi linear condition, we still see the head of the beam stabilising after some time. The off-axis
case has a larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the growth levels
off after the first few metres and stays constant for the remainder of the plasma accelerator section (figure
\ref{Fig:BeamEmitt}).

@@

For accelerator applications in general, and also for Run 2 of the AWAKE experiment, it is desirable to maximise the
charge that can be successfully accelerated in the witness bunch. However, a longitudinally constant accelerating
field depends on both the position and the charge density of the witness beam \cite{katsouleas:1987, tzoufras:2009}.

@@

A higher charge beam will generate a non-linear wakefield closer to the head of the bunch, which in turns ensures that a
smaller portion of the beam experiences the quasi-linear region with emittance growth. As figure \ref{Fig:BeamQAbs}
illustrates, a higher charge also results in a lower energy gain.

>>>>

\title{Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield}

@@

\begin{abstract}
We investigate beam loading and emittance preservation for a high-charge electron beam being accelerated in quasi-linear
plasma wakefield driven by a short proton beam. The structure of the wakefield is similar to that of a long, modulated
proton beam, such as the one being studied by AWAKE.  We show that by exploiting two well known effects, full blow out
of plasma electrons by the accelerated beam, and beam loading of the wake field, the electron beam can be accelerated
without significant emittance growth.
\end{abstract}

@@

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativist beam travelling through a plasma
excites a strong longitudinal e-field that can be loaded by a trailing witness beam. With optimal beam loading, increase
in energy spread can be kept to a minimum. Acceleration of an electron withness beam by an electron drive beam has been
demonstrated experimentally \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} in the past. AWAKE at CERN is a proof
of concept proton driven plasma wakefield accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, the non-linear transverse wakefields cause emittance growth of the accelerated
beam. A finite length beam will also see a transversely varying accelerating field causing increasing energy spread
\cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, the drive beam sees a constant transverse field
due to a bubble forming in the plasma behind the drive beam. The bubble is formed by the transverse oscillations of the
plasma electrons in the wake of the drive beam, forming a sheet around an evacuated area. The ions, assumed stationary,
form an ion channel creating a focusing force that varies linearly with radius. This
produces an axially symmetric focusing force \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

The energy carried by electron drive bunches used in previous experiments were typically small, on the order of
$100\unit{J}$, and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. De-phasing of the
witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
beam, limits the effective length of a plasma accelerator stage. Proton drive beams available at CERN carry
significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for longer accelerator
stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications.
This issue is resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of
the buckets in the modulated structure. This self-modulation instability is caused by the transverse fields generated by
the beam acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation
frequency is close to that of the plasma, and leaves a train of proton bunches along the beam axis with a surrounding
halo.

% ******************************************************************************************************************** %
\subsection[\label{S:I:AWAKE}]{AWAKE Run 2}
% ******************************************************************************************************************** %

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver, and a nominal plasma density $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron beam. The study presented here is for Run 2 \cite{adli:2016}, which aims to demonstrate
acceleration of a short electron beam to high energy with a minimal increase in emittance and energy spread.

@@

The preliminary design of the second run proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without
the electron beam present. The electron witness beam will be injected into the modulated proton beam before stage two,
where acceleration will occur.
%~ As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
%~ keep this as short as possible.

With this experimental design the self-modulated proton beam does not produce a non-linear wakefield, and not all plasma
electrons are therefore evacuated. The result is that the focusing force will not increase linearly with radius. The
small size of the accelerating structure produced by the self-modulated beam also put constraints on the size of the
electron witness beam, and thus its charge and current, if we wish to prevent large energy spread. By matching the
transverse size to a practical emittance and the plasma density, we prevent large amplitude oscillations which may cause
additional growth in energy spread as well as in emittance.

The key idea in this paper is that while the proton beam wake is not fully blown-out leading to non-linear focusing
forces, the electron beam, if intense enough, may load the field and at the same time create its own ion bubble. As will
be shown, exploiting these two well known effect, the electron beam can be accelerated, for long distances, without
significant emittance growth.

%~ The beam matching relation is given by
%~ \begin{align}
    %~ \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
%~ \end{align}
%~ where $\lambda_{p}$ is the plasma wavelength. For a reasonable initial emittance of $2\unit{\mu m}$, this corresponds to
%~ a $\sigma_{r}$ of $5.25\unit{\mu m}$ -- a very narrow bunch compared to the drive beam $\sigma_{r} = 200\unit{\mu m}$.
%~ The charge density of this compact witness bunch reaches that of the plasma density at only a few $\unit{pC}$, but
%~ reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the witness bunch produces its
%~ own non-linear wakefield already at the head of the bunch. The majority of the electrons within it will therefore see a
%~ linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations \cite{berglyd_olsen:2015, berglyd_olsen:2016}.

%~ In order to evaluate the quality of the accelerated bunch it was also necessary to study emittance evolution. However,
%~ full PIC codes are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect''
%~ \cite{godfrey:1974}. This is a know issue with the Yee EMF solver which causes the phase velocity of electromagnetic
%~ fields to be lower than $c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing
%~ the Yee solver with a solver designed by Lehe \cite{lehe:2013}. Our simulations showed that this numerical effect is
%~ still prominent in the high density regions of our very compact electron bunch, making it difficult to distinguish
%~ emittance growth from the physics from that originating from numerical error.

In order to study the witness beam emittance evolution we turned to the recently released open source version of
QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}.
Since QuickPIC is a quasi-static code, it does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, it is well suited to study emittance preservation.

@@

In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}). When the
initial SPS beam containing $3\nexp{11}$ protons \cite{gschwendtner:2016} enters the second plasma cell the peak $E_{z}$
field is expected to be $500-600\funit{MV}{m}$. As the baseline AWAKE drive beam current is insufficient to reach the
non-linear regime and produce a plasma bubble, the plasma electrons are only depleted to around $65\%$ of nominal plasma
density at the point where we inject the electron beam \cite{awake_collaboration:2016}. These conditions are replicated
in our simulations using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and $7\unit{kA}$.

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{r} = 200\unit{\mu m}$. We
set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. In order to create a stable environment to study the
evolution of the witness beam we prevented the proton beam from evolving radially by increasing the particle mass by a
factor of $1\mexp{6}$. We are not considering the evolution of the proton beam itself in this study.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. The relative drift causing the de-phasing $\Delta\xi \propto \gamma^{-2}$. By simply matching the
initial $\gamma$ values we eliminate this effect entirely, but a lower initial value is likely to be sufficient
\cite{berglyd_olsen:2015}.

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{r}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. This is a very narrow beam compared to the drive beam
$\sigma_{r} = 200\unit{\mu m}$. The charge density of this compact witness beam reaches that of the plasma density at
only a few $\unit{pC}$, but reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the
witness beam produces its own non-linear wakefield already at the head of the beam. The majority of the electrons
within it will therefore see a linear focusing force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam. We used a transverse grid cell size of $1.17\unit{\mu m}$, and $2.34\unit{\mu m}$ for
the longitudinal grid cells. The witness beam was simulated with $16.8\unit{M}$ non-weighted particles.

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent wakefields throughout the plasma stage. The $E_{z}$-field generated by the proton drive beam is seen as
the blue line in figure \ref{Fig:BeamLoading}, shown with and without the electron beam present. With a proton beam
density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near non-linear \cite{rosenzweig:2010}. There is
some depletion of plasma electrons, but not enough to form a bubble. The dashed green line in the lower part of figure
\ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of $67\%$, close to what we see in full
scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

The witness beam generates its own wakefield, which in the accelerating phase partially cancels out the $E_{z}$ field
generated by the drive beam. For a finite length beam this causes the particles in the tail of the witness beam to be
accelerated less than those at the front resulting in higher energy spread in exchange for higher energy transfer
\cite{van_der_meer:1985}. Since $E_{z} \propto \rho_{b}$, a beam profile $\rho_{b}(\xi)$ that perfectly flattens the
accelerating field within the witness beam will reduces energy spread to zero. The ideal shape is triangular or
trapezoidal with the peak density in the direction of the beam velocity \cite{katsouleas:1987, tzoufras:2009}.

@@

For Gaussian beams a reasonable flat field, and consequently low energy spread, can be achieved by controlling the
charge density to achieve a similar effect. This, however, puts the front of the witness beam outside of the ideal
region while the middle bulk of the beam has a trapezoidal-like shape. In our case, with a matched beam to the plasma
density defined by equation \ref{EQ:Matched}, we are limited on how much total charge we can accelerate without reaching
charge densities that will overload the field. The test case shown in figures \ref{Fig:BeamLoading} and
\ref{Fig:PlasmaDenTWake} has a total charge of $100\unit{pC}$ with a $\sigma_{r}=5.23\unit{\mu m}$ matching an initial
normalised emittance $\epsilon_{0} = 2\unit{\mu m}$.

@@

Our configuration also shows some robustness to small electron beam transverse offsets as the bubble generated by the
witness beam wakefield follows the beam itself. The head of the beam does not benefit from this effect, and we see
some defocusing in this region. This effect is likely to be greater for larger offsets as the beam oscillates around
the axis of the drive beam wakefield (see right hand plots of figure \ref{Fig:BeamFilament}). However, since the proton
wakefield creates a quasi linear condition, we still see the head of the beam stabilising after some time. The off-axis
case has a larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the growth levels
off after the first few metres and stays constant for the remainder of the plasma accelerator section (figure
\ref{Fig:BeamEmitt}).

@@

For accelerator applications in general, and also for Run 2 of the AWAKE experiment, it is desirable to maximise the
charge that can be successfully accelerated in the witness beam. However, a longitudinally constant accelerating
field depends on both the position and the charge density of the witness beam \cite{katsouleas:1987, tzoufras:2009}.

@@

A higher charge beam will generate a non-linear wakefield closer to the head of the beam, which in turns ensures that a
smaller portion of the beam experiences the quasi-linear region with emittance growth. As figure \ref{Fig:BeamQAbs}
illustrates, a higher charge also results in a lower energy gain.
